Automating Political Bias Prediction

### 2016-06-12 manuscript/poltext.tex
Some deletions.

\section{Introduction}
\label{sec:intro}
%
Analysis and classifications of political text is and has been a very important tool to generate political science data (Benoit et al. 2015). Traditionally such classifications are done by experts, who read and label the text of interest (see for example the Manifesto Project (Budge et al. 2001; Klingemann et al. 2006), the Comparative Agendas Project or Poltext (Pétry/Duval 2015)). This is, however, a very time consuming task and thus sets various limits to the possible amount of data that a few experts can analyze. The growing field of automated text analysis, that allows the analysis of much more text in less time, is therefore of great interest to political scientists. Additionally automated text analyses allow for a more objective and replicable analysis of political text then human coders could achieve (Benoit et al. 2009)

@@

\subsection{Data}
Annotated political text data was obtained from two sources: a) the discussions and speeches held in the German parliament ({\em Bundestag}) and b) all manifesto texts of parties winning seats in the election to the German parliament in the current 18th and the last, 17th, legislative period. For the 18th period we also added the manifesto data of three additional parties that did not win seats but nonetheless played a very important role in the political discourse in the run-up to the election.

@@

\paragraph{Party manifesto data}
The party manifesto text was taken from the Manifesto Corpus (add citation). The Corpus is accessible through an publicly available API published by the {\em Manifesto Project} \cite{manifesto}. The data released in this project mainly comprises the complete manifestos of all parties that have won seats at a national election. Each quasi-sentence \footnote{A quasi-sentence has the length of an argument. It is never longer than one sentence.} is annotated with one of 56 policy goals. Examples for the policy categories are {\em welfare state expansion, welfare state limitation, democracy, equality}; for a complete list and detailed explanations on how the annotators were instructed see \cite{leftright}. In total the 12 manifestos from the 17th and 18th legislative period have been divided into 29451 quasi-sentences. Each observation has been added with two labels: the party affiliation and the manually assigned policy goal aimed at in each specific statement. The party affiliation was used to evaluate the party evaluation classifiers trained on the parliament speeches. For this purpose the data acquisition was constrained to only those parties that were elected into the parliament. For the analyses based on policy issues all parties were considered, also those three that did not make it into parliament.

The length of each annotated statement in the party manifestos was rather short. The median length was 95 characters or 12 words. \footnote{The longest statement was 522 characters long, the 25\%/50\%/75\% percentiles were 63/95/135 characters. Measured in words the longest data point was 65 words and the 25\%/50\%/75\% percentiles were 8/12/17 words, respectively.} This can be considered as a very valuable property of the data set, because it allows a fine grained resolution of party manifestos and it a good test for media data, as some data like tweets, facebook texts etc. are often very short, too. However for a classifier (as well as for humans) such short sentences can be rather difficult to classify. Benoit et al. (2015), for example, found in an experiment where experts where differentiating whether specific sentences were dealing with economic or social policy only about 35\% agreement between all expert coders. In the Manifesto Project the human coders do not code these statements separately but within context. In order to obtain less 'noisy' data points from each party -- for the party affiliation task only -- all statements were aggregated into eight policy domains. Each of the 56 policy categories from the Manifesto Project belongs to one of seven policy domains. These were used to aggregate the data into topics. An eight topic is generated for all statements not belonging to any one the 56 policy issues: {\em External Relations, Freedom and Democracy, Political System, Economy, Welfare and Quality of Life, Fabric of Society, Social Groups} and a topic {\em undefined}, for a complete list see also \cite{leftright}. Most party manifestos covered all eight of them, some party manifestos in the 17th Bundestag only covered seven.

\subsection{Bag-of-Words Vectorization}\label{sec:bow-vectorization}
First each data set was segmented into semantic units; in the case of the party manifesto data semantic units were the sentences or sentence parts associated with one of the 56 policy issues, in the case of parliament discussions this were the speeches. Parliament speeches were often interrupted; in this case each uninterrupted part of a speech was considered a semantic unit. Strings of each semantic unit were tokenised and transformed into bag-of-word vectors as implemented in scikit-learn \cite{scikit-learn}. The general idea of bag-of-words vectors is to simply count occurrences of words (or word sequences, also called {\em n-grams}) for each data point. A data point is usually a document, here it is the semantic units of parliament speeches and manifesto sentences, respectively. The text of each semantic unit is transformed into a vector $\vec{x}\in\mathds{R}^d$ where $d$ is the size of the dictionary; the $w$th entry of $\vec{x}$ contains the (normalized) count of the $w$th word (or sequence of words) in our dictionary. Several options for vectorizing the speeches were tried, including term-frequency-inverse-document-frequency normalisation, n-gram patterns up to size $n=3$ and several cutoffs for discarding too frequent and too infrequent words. All of these hyperparameters were subjected to hyperparameter optimization as explained in \autoref{sec:crossvalidation}.

\section{Classification Model and Training Procedure}\label{sec:model}
Bag-of-words feature vectors were used to train a multinomial logistic regression model. Let $y\in\{1,2,\dots,K\}$ be the true  label, where $K$ is the total number of labels and $\vec{W}=[\vec{w}_1,\dots,\vec{w}_K]\in\R^{d\times K}$ is the concatenation of the weight vectors $\vec{w}_k$ associated with the $k$th party then
\begin{eqnarray}\label{eq:logreg_multiclass}
p(y=k|\vec{x},\vec{W}) = &\frac{e^{z_k}}{\sum_{j=1}^K e^{z_j}} \qquad \textrm{with }  z_k=&\vec{w}_k^{\top}\vec{x} \\\nonumber
\end{eqnarray}
%
We estimated $\vec{W}$ using quasi-newton gradient descent. The optimization function was obtained by adding a penalization term to the negative log-likelihood of the multinomial logistic regression objective and the optimization hence found the $\vec{W}$ that minimized
\begin{equation}\label{eq:objective}
L(\vec{W}, \vec{x}, \gamma) = - \log{\frac{e^{z_k}}{\sum_{j=1}^K e^{z_j}}}+ \gamma \| \vec{W} \|_{F}
\end{equation}
Where $\|~\|_F$ denotes the Frobenius Norm and $\gamma$ is a regularization parameter controlling the complexity of the model.
The regularization parameter was optimized on a log-scaled grid from $10^{-4,\dots,4}$. The performance of the model was optimized using the classification accuracy, but we also report all other standard measures, precision ($TP / (FP + TP$), recall ($TP / (TP + FN)$) and f1-score ($2\times (Prec. \times Rec) / (Prec + Rec.)$). \\

@@

\subsection{Optimisation of Model Parameters}\label{sec:crossvalidation}
The model pipeline contained a number of  hyperparameters that were optimised using cross-validation.
We first split the training data into a training data set that was used for optimisation of hyperparameters and an held-out test data set for evaluating how well the model performs on in-domain data; wherever possible the generalisation performance of the models was also evaluated on out-of domain data. Hyperparameters were optimised using grid search and 3-fold cross-validation within the training set only: A cross-validation split was made to obtain train/test data for the grid search and for each setting of hyperparameters the entire pipeline was trained and evaluated -- no data from the in-domain evaluation data or the out-of-domain evaluation data were used for hyperparameter optimisation. For the best setting of all hyperparameters the pipeline was trained again on all training data and evaluated on the evaluation data sets. For party affiliation prediction and government membership prediction the training and test set were 90\% and 10\%, respectively, of all data in a given legislative period. Out-of-domain evaluation data were the texts from party manifestos. For the policy issue prediction setting there was no out-of-domain evaluation data, so all labeled manifesto sentences in both legislative periods were split into a training and evaluation set of 90\% (train) and 10\% (evaluation).

@@

The following section gives an overview of the results for all political bias prediction tasks. Some interpretations of the results are highlighted and a web application of the models is presented at the end of the section.

\subsection{Predicting political party affiliation}
The results for the political party affiliation prediction on in-domain data (held-out parliamentary speech text) for the 17th and 18th Bundestag are listed in \autoref{tab:results_in-domain}.
Shown are the evaluation results for in-domain data.

When predicting party affiliation on text data from the same domain that was used for training the model, average precision and recall values of above 0.6 are obtained. These results are comparable to those of \cite{Hirst2014} who report a classification accuracy of 0.61 on a five class problem of prediction party affiliation in the European parliament; the accuracy for the 17th Bundestag is 0.63, results of the 18th Bundestag are difficult to compare as the number of parties is four and the legislation period is not finished yet.
But as the main purpose of this paper is to generate a model that can be used to predict political bias on new data, like media outtakes, we are interested in the question if the model also produces satisfying results on out-of-domain data. We are therefore also predicting party affiliation on party manifesto text. For out-of domain data the models yield significantly lower precision and recall values between 0.3 and 0.4 \autoref{tab:results_out-of-domain}. This drop in out of domain prediction accuracy is in line with previous findings \cite{Yu2008}.
A main factor that made the prediction on the out-of-domain prediction task particularly difficult is the short length of the strings to be classified, see also \autoref{sec:data}. In order to investigate whether this low out-of-domain prediction performance was due the domain difference (parliament speech vs manifesto data) or due to the short length of the data points, the manifesto data was aggregated based on the political domain into which each quasi-sentence was coded, see \autoref{sec:data}. The topic level results are shown in \autoref{tab:results_topic} and demonstrate that when the texts to be classified are sufficiently long and the word count statistics are sufficiently dense the classification performance on out of domain data can achieve in the case of some parties reliable precision and recall values close to 1.0. This increase is in line with previous findings on the influence of text length on political bias prediction accuracy \cite{Hirst2014}.

\begin{table}[t]
\caption{
\label{tab:results_in-domain}
Classification performance on the party affiliation prediction problem for data from the 17th and 18th legislative period on in-domain data. $N$ denotes number of data points in the evaluation set.
}
\begin{center}
\begin{tabular}{lcccccccc}
& \multicolumn{4}{c}{\bf Held-out parliament speeches} & \multicolumn{4}{c}{\bf Held-out parliament speeches}\\
& \multicolumn{4}{c}{\bf 17th legislative period} & \multicolumn{4}{c}{\bf 18th legislative period}\\
    &         precision    &recall &  f1-score  & N    &         precision    &recall &  f1-score  & N\\
\hline \hline
       cducsu   &    0.62  &    0.81  &    0.70  &     706&0.66    &  0.82   &   0.73    &  456\\
        fdp    &   0.70   &   0.37  &    0.49    &   331&   &      &     &   0\\
     gruene &      0.59  &    0.40   &   0.48   &    298&0.68  &    0.54   &   0.60 &     173\\
      linke    &   0.71   &   0.61  &    0.65    &   338&0.77   &   0.58   &   0.66    &  173\\
        spd   &    0.60   &   0.69  &    0.65   &    606&0.60   &   0.54   &   0.57  &    330\\
\hline
avg / total &      0.64   &   0.63   &   0.62    &  2279 &0.66    &  0.66 &     0.30   &  1132
%
\end{tabular}
\end{center}
\end{table}

\begin{table}[t]
\caption{
\label{tab:results_out-of-domain}
Classification performance on the party affiliation prediction problem for data from the 18th legislative period on test set and evaluation set, respectively.  Predictions on the manifesto data was done on {\bf sentence level}.
}
\begin{center}
\begin{tabular}{lcccccccc}
& \multicolumn{4}{c}{\bf Party Manifestos} & \multicolumn{4}{c}{\bf Party Manifestos}\\
& \multicolumn{4}{c}{\bf 17th legislative period} & \multicolumn{4}{c}{\bf 18th legislative period}\\
    &         precision    &recall &  f1-score  & N    &         precision    &recall &  f1-score  & N\\
\hline \hline
    cducsu    &   0.26   &   0.58   &   0.36    &   2030 & 0.32  &    0.64  &    0.43    &  2983\\
    fdp    &   0.38   &   0.28   &   0.33    &   2319 &   &      &        &  0\\
     gruene   &    0.47    &  0.20   &   0.28    &  3747   &0.59   &   0.15   &   0.23   &   5674\\
      linke     &  0.30  &    0.47    &  0.37    &   1701 & 0.36   &   0.48   &   0.41   &   2555\\
        spd     &  0.26  &    0.16   &   0.20    &   2278 & 0.26 &     0.31   &   0.28     & 2989\\
\hline
avg / total    &   0.35  &    0.31  &    0.30   &   12075 &  0.42 &     0.34  &    0.32&     14201\\
%
\end{tabular}
\end{center}

@@

\begin{table}[t]
\caption{
\label{tab:results_topic}
{\bf Topic level classification performance} on the party affiliation prediction problem for data from the evaluation set (manifesto texts) of the 17th legislative period. In contrast to single sentence level predictions (see \autoref{tab:results_17}, \autoref{tab:results_18}, \autoref{tab:confusion} for results and \autoref{sec:data} for topic definitions) the predictions made on topic level are reliable in many cases. Note that all manifesto topics of the green party in the 18th Bundestag are predicted to be from the parties of the governing coalition, CDU/CSU or SPD.}
\begin{center}
\begin{tabular}{lcccc}
& \multicolumn{4}{c}{\bf Party Manifestos} \\
& \multicolumn{4}{c}{\bf 17th Bundestag} \\
    &         precision    &recall &  f1-score  & N  \\
    \hline
        \hline
cducsu     &  0.64  &    1.00  &    0.78    &     7\\
       fdp    &   1.00    &  1.00    &  1.00    &     7\\
    gruene  &     1.00  &    0.86  &    0.92    &     7\\
     linke    &   1.00   &   1.00     & 1.00    &     7\\
       spd   &    0.80   &   0.50    &  0.62     &    8\\
    \hline
    avg / total  &     0.88   &   0.86   &   0.86  &      36\\
\end{tabular}
\quad
\begin{tabular}{lcccc}
& \multicolumn{4}{c}{\bf Party Manifestos} \\
& \multicolumn{4}{c}{\bf 18th Bundestag} \\
    &         precision    &recall &  f1-score  & N  \\
    \hline
    \hline
  cducsu     &  0.50   &   1.00  &    0.67     &    8\\
       gruene  &     0.00   &   0.00   &   0.00  &       8\\
      linke     &  1.00   &   0.88   &   0.93  &       8\\
        spd     &  0.56  &    0.62  &    0.59   &      8\\
\hline
avg / total   &    0.51 &     0.62   &   0.55   &     32\\
\end{tabular}
\end{center}

@@

In order to investigate the errors conducted by the models confusion matrices were extracted for the predictions on the out-of-domain evaluation data for sentence level predictions (see \autoref{tab:confusion}) as well as topic level predictions (see \autoref{tab:confusion_topic}). One example illustrates that the mistakes the model makes can be associated with changes in the party policy. The green party has been promoting policies for renewable energy and against nuclear energy in their manifestos prior to both legislative periods. Yet the statements of the green party are more often predicted to be from the government parties than from the party that originally promoted these green ideas, reflecting the trend that during the legislative period governing parties took over policies from the green party. This effect is even more pronounced in the topic level predictions: a model trained on data from the 18th Bundestag predicts all manifesto content of the green party to be from one of the parties of the governing coalition, CDU/CSU or SPD. \\

% Confusion matrix sentence level
\begin{table}[t]\label{tab:conf_mat_four_class}
\caption{\label{tab:confusion} {\bf Confusion matrices (sentence level)} for predictions on evaluation data (party manifestos); classifiers were trained on parliament speeches for the 17th legislative period (left) and 18th legislative period (right); the most prominent effect is the high likelihood for a party to be taken as the strongest, governing party, cdu/csu. This can be interpreted as a change in policies of the conservative party cdu/csu towards the policies of the green party.}
\begin{tabular}{lccccccc}
 \multicolumn{8}{c}{\bf Party Manifestos}\\
 \multicolumn{8}{c}{\bf 17th Bundestag}\\
 \\
&&& \multicolumn{5}{c}{Predicted}\\
&&& cducsu & fdp& gruene& linke& spd\\
\hline
\multirow{5}{*}{\rotatebox{90}{\pbox{3cm}{\centering True}}}& &cducsu &1186 &289& 178& 198& 179\\
&&fdp &882& 658& 236& 329& 214\\
&&gruene &1174& 404& 764& 941& 464\\
&&linke &388& 92& 214& 806& 201\\
&&spd &999& 268& 240& 398& 373\\
\end{tabular}
\hfill
\begin{tabular}{lcccccc}
 \multicolumn{7}{c}{\bf Party Manifestos}\\
 \multicolumn{7}{c}{\bf 18th Bundestag}\\
 \vspace{1em}\\
&&& \multicolumn{4}{c}{Predicted}\\
&&& cducsu & gruene& linke& spd\\
\hline
\multirow{4}{*}{\rotatebox{90}{\pbox{4.7cm}{\centering True}}}&&cducsu&1912& 156& 331& 584\\
&&gruene&2092& 827& 1311& 1444\\
&&linke&596& 186& 1216& 557\\
&&spd&1284& 226& 563& 916\\
\end{tabular}
\vspace{1em}
\end{table}

% Confusion matrix topic level
\begin{table}[t]\label{tab:conf_mat_four_class}
\caption{\label{tab:confusion_topic} {\bf Confusion matrices (topic level)} for predictions on evaluation data (party manifestos) for classifiers trained on parliament speeches for the 17th legislative period (left) and 18th legislative period (right).}
\begin{tabular}{lccccccc}
 \multicolumn{8}{c}{\bf Party Manifestos}\\
 \multicolumn{8}{c}{\bf 17th Bundestag}\\
 \\
&&& \multicolumn{5}{c}{Predicted}\\
&&& cducsu & fdp& gruene& linke& spd\\
\hline
\multirow{5}{*}{\rotatebox{90}{\pbox{3cm}{\centering True}}} &&cducsu &7& 0& 0& 0& 0\\
&&fdp&0& 7& 0& 0& 0\\
&&gruene&0& 0& 6& 0& 1\\
&&linke&0& 0& 0& 7& 0\\
&&spd&4& 0& 0& 0& 4\\
\end{tabular}
\hfill
\begin{tabular}{lcccccc}
 \multicolumn{7}{c}{\bf Party Manifestos}\\
 \multicolumn{7}{c}{\bf 18th Bundestag}\\
 \vspace{1em}\\
&&& \multicolumn{4}{c}{Predicted}\\
&&& cducsu & gruene& linke& spd\\
\hline
\multirow{4}{*}{\rotatebox{90}{\pbox{4.7cm}{\centering True}}}&&cducsu&8& 0& 0& 0\\
&&gruene&4& 0& 0& 4\\
&&linke&1& 0& 7& 0\\
&&spd&3& 0& 0& 5\\
\end{tabular}
\vspace{1em}\\
\end{table}

>>>>

\section{Introduction}
\label{sec:intro}
%
Analysis and classifications of political text is and has been a very important tool to generate political science data (Benoit et al. 2015). Traditionally such classifications are done by experts, who read and label the text of interest.\footnote{See for example the Manifesto Project (Budge et al. 2001; Klingemann et al. 2006), the Comparative Agendas Project or Poltext (Pétry/Duval 2015).} This is, however, a very time consuming task and thus sets various limits to the possible amount of data that a few experts can analyze. The growing field of automated text analysis, that allows the analysis of much more text in less time, is therefore of great interest to political scientists. Additionally automated text analyses allow for a more objective and replicable analysis of political text then human coders could achieve (Benoit et al. 2009)

@@

\subsection{Data}
Annotated political text data was obtained from two sources: a) the plenary debates held in the German parliament ({\em Bundestag}) and b) all manifesto texts of parties winning seats in the election to the German parliament in the current 18th and the last, 17th, legislative period.

%For the 18th period we also added the manifesto data of three additional parties that did not win seats but nonetheless played a very important role in the political discourse in the run-up to the election. Wollen wir die einfach rauslassen? Käme ja eh nur in einem kleinen Teil der Analyse vor und macht die Beschreibung an mehreren Stellen kürzer.

@@

\paragraph{Party manifesto data}
The party manifesto text was taken from the Manifesto Corpus (add citation). The Corpus is accessible through an publicly available API published by the {\em Manifesto Project} \cite{manifesto}. The data released in this project mainly comprises the complete manifestos of all parties that have won seats at a national election. Each quasi-sentence\footnote{A quasi-sentence has the length of an argument. It is never longer than one sentence.} is annotated with one of 56 policy issue categories. Examples for the policy categories are {\em welfare state expansion, welfare state limitation, democracy, equality}; for a complete list and detailed explanations on how the annotators were instructed see \cite{leftright}. In total the 9 manifestos from the 17th and 18th legislative period have been divided into 29451 quasi-sentences. Each observation has been added with two labels: the party affiliation and the manually assigned policy issue aimed at in each specific statement.

%The party affiliation was used to evaluate the party evaluation classifiers trained on the parliament speeches. For this purpose the data acquisition was constrained to only those parties that were elected into the parliament. For the analyses based on policy issues all parties were considered, also those three that did not make it into parliament.

The length of each annotated statement in the party manifestos is rather short. The median length is 95 characters or 12 words.\footnote{The longest statement is 522 characters (65 words) long, the 25\%/50\%/75\% percentiles are 63/95/135 characters or 8/12/17 words, respectively.} This can be considered as a very valuable property of the data set, because it allows a fine grained resolution of party manifestos and is a good test for media data, as some data like tweets, facebook texts etc. are often very short, too. However for a classifier (as well as for humans) such short sentences can be rather difficult to classify. Benoit et al. (2015), for example, found in an experiment where experts where differentiating whether specific sentences were dealing with economic or social policy only about 35\% agreement between all expert coders. In the Manifesto Project the human coders therefore do not code these statements separately but within context. In order to obtain less 'noisy' data points from each party -- for the party affiliation task only -- all statements were aggregated into eight policy domains using the 56policy categories from the Manifesto Project. Within the Manifesto Project category scheme each of these belongs to one of seven policy domains. These were used to aggregate the data into the following topics: {\em External Relations, Freedom and Democracy, Political System, Economy, Welfare and Quality of Life, Fabric of Society, Social Groups}. An eight topic is generated for all statements not belonging to any one the 56 policy issues, for a complete list see also \cite{leftright}. Most party manifestos covered all eight topics, some party manifestos in the 17th Bundestag only covered seven.

\subsection{Bag-of-Words Vectorization}\label{sec:bow-vectorization}
First each data set was segmented into semantic units; in the case of the party manifesto data the semantic units were the quasi-sentences associated with one of the 56 policy issues, in the case of parliament discussions this were the speeches. Parliament speeches were often interrupted; in this case each uninterrupted part of a speech was considered a semantic unit. Strings of each semantic unit were tokenised and transformed into bag-of-word vectors as implemented in scikit-learn \cite{scikit-learn}. The general idea of bag-of-words vectors is to simply count occurrences of words (or word sequences, also called {\em n-grams}) for each data point. A data point is usually a document, here it is the semantic units of parliament speeches and manifesto sentences, respectively. The text of each semantic unit is transformed into a vector $\vec{x}\in\mathds{R}^d$ where $d$ is the size of the dictionary; the $w$th entry of $\vec{x}$ contains the (normalized) count of the $w$th word (or sequence of words) in our dictionary. Several options for vectorizing the speeches were tried, including term-frequency-inverse-document-frequency normalisation, n-gram patterns up to size $n=3$ and several cutoffs for discarding too frequent and too infrequent words. All of these hyperparameters were subjected to hyperparameter optimization as explained in \autoref{sec:crossvalidation}.

\section{Classification Model and Training}\label{sec:model}
Bag-of-words feature vectors were used to train a multinomial logistic regression model. Let $y\in\{1,2,\dots,K\}$ be the true  label, where $K$ is the total number of labels and $\vec{W}=[\vec{w}_1,\dots,\vec{w}_K]\in\R^{d\times K}$ is the concatenation of the weight vectors $\vec{w}_k$ associated with the $k$th party then
\begin{eqnarray}\label{eq:logreg_multiclass}
p(y=k|\vec{x},\vec{W}) = &\frac{e^{z_k}}{\sum_{j=1}^K e^{z_j}} \qquad \textrm{with }  z_k=&\vec{w}_k^{\top}\vec{x} \\\nonumber
\end{eqnarray}
The performance of the model was optimized using the classification accuracy, but we also report all other standard measures, precision ($TP / (FP + TP$), recall ($TP / (TP + FN)$) and f1-score ($2\times (Prec. \times Rec) / (Prec + Rec.)$). \\

@@

\subsection{Optimisation of Model Parameters}\label{sec:crossvalidation}
The model pipeline contained a number of  hyperparameters that were optimised using cross-validation.
We first split the training data into a training data set that was used for optimisation of hyperparameters and an held-out test data set for evaluating how well the model performs on in-domain data; wherever possible the generalisation performance of the models was also evaluated on out-of domain data. Hyperparameters were optimised using grid search and 3-fold cross-validation within the training set only: A cross-validation split was made to obtain train/test data for the grid search and for each setting of hyperparameters the entire pipeline was trained and evaluated -- no data from the in-domain evaluation data or the out-of-domain evaluation data were used for hyperparameter optimisation. For the best setting of all hyperparameters the pipeline was trained again on all training data and evaluated on the evaluation data sets. For party affiliation prediction and government membership prediction the training and test set were 90\% and 10\%, respectively, of all data in a given legislative period. Out-of-domain evaluation data were the texts from party manifestos. For the policy issue prediction setting there was no out-of-domain evaluation data, so all labeled manifesto sentences in both legislative periods were split into a training and evaluation set of 90\% (train) and 10\% (evaluation).
% Falls wir die Medien nochreinnehmen, muss das bei out-of-domain auch noch erwähnt werden.

@@

The following section gives an overview of the results for all political bias prediction tasks. Some interpretations of the results are highlighted and a web application of the models is presented at the end of the section. Because of space restrictions, we will only report results fromt he 17th legislative period here. Results are very similar for the 18th legislative period and can be accessed via github.

\subsection{Predicting political party affiliation}
The evaluation results for the political party affiliation prediction on in-domain data (held-out parliamentary speech text) for the 17th Bundestag are listed in \autoref{tab:results_in-domain}.

When predicting party affiliation on text data from the same domain that was used for training the model, average precision and recall values of above 0.6 are obtained. These results are comparable to those of \cite{Hirst2014} who report a classification accuracy of 0.61 on a five class problem predicting party affiliation in the European parliament; the accuracy for the 17th Bundestag is 0.63, results of the 18th Bundestag are difficult to compare as the number of parties is four and the legislation period is not finished yet.

But as the main purpose of this paper is to generate a model that can be used to predict political bias on new data, like media texts, we are interested in the question if the model also produces satisfying results on out-of-domain data. We are therefore also predicting party affiliation on party manifesto text. For out-of domain data the models yield significantly lower precision and recall values between 0.3 and 0.4 \autoref{tab:results_out-of-domain}. This drop in out of domain prediction accuracy is in line with previous findings \cite{Yu2008}.
A main factor that made the prediction on the out-of-domain prediction task particularly difficult is the short length of the strings to be classified, see also \autoref{sec:data}. In order to investigate whether this low out-of-domain prediction performance was due to the domain difference (parliament speech vs manifesto data) or due to the short length of the data points, the manifesto data was aggregated into the previously described party domains, see \autoref{sec:data}. The topic level results are shown in \autoref{tab:results_topic} and demonstrate that when the texts to be classified are sufficiently long and the word count statistics are sufficiently dense the classification performance on out-of-domain data can - at least in some cases - achieve reliable precision and recall values close to 1.0. This increase is in line with previous findings on the influence of text length on political bias prediction accuracy \cite{Hirst2014}.

\begin{table}[t]
\caption{
\label{tab:results_in-domain}
Classification performance on the party affiliation prediction problem for data from the 17th legislative period on in-domain data. $N$ denotes number of data points in the evaluation set.
}
\begin{center}
\begin{tabular}{lcccc}
    &         precision    &recall &  f1-score  & N  \\
\hline \hline
       cducsu   &    0.62  &    0.81  &    0.70  &     706\\
        fdp    &   0.70   &   0.37  &    0.49    &   331\\
     gruene &      0.59  &    0.40   &   0.48   &    298\\
      linke    &   0.71   &   0.61  &    0.65    &   338\\
        spd   &    0.60   &   0.69  &    0.65   &    606\\
\hline
avg / total &      0.64   &   0.63   &   0.62    &  2279 
%
\end{tabular}
\end{center}
\end{table}

\begin{table}[t]
\caption{
\label{tab:results_out-of-domain}
Classification performance on the party affiliation prediction problem for data from the 17th legislative period on out-of-domain data.  Predictions are done on the {\bf sentence level}.
}
\begin{center}
\begin{tabular}{lcccc}
    &         precision    &recall &  f1-score  & N  \\
\hline \hline
    cducsu    &   0.26   &   0.58   &   0.36    &   2030 \\
    fdp    &   0.38   &   0.28   &   0.33    &   2319 \\
     gruene   &    0.47    &  0.20   &   0.28    &  3747\\
      linke     &  0.30  &    0.47    &  0.37    &   1701\\
        spd     &  0.26  &    0.16   &   0.20    &   2278\\
\hline
avg / total    &   0.35  &    0.31  &    0.30   &   12075\\
%
\end{tabular}
\end{center}

@@

\begin{table}[t]
\caption{
\label{tab:results_topic}
{\bf Topic level classification performance} on the party affiliation prediction problem for data from the evaluation set (manifesto texts) of the 17th legislative period. In contrast to single sentence level predictions (see \autoref{tab:results_17}, \autoref{tab:results_18}, \autoref{tab:confusion} for results and \autoref{sec:data} for topic definitions) the predictions made on topic level are reliable in many cases. Note that all manifesto topics of the green party in the 18th Bundestag are predicted to be from the parties of the governing coalition, CDU/CSU or SPD.}
\begin{center}
\begin{tabular}{lcccc}
& \multicolumn{4}{c}{\bf Party Manifestos} \\
& \multicolumn{4}{c}{\bf 17th Bundestag} \\
    &         precision    &recall &  f1-score  & N  \\
    \hline
        \hline
cducsu     &  0.64  &    1.00  &    0.78    &     7\\
       fdp    &   1.00    &  1.00    &  1.00    &     7\\
    gruene  &     1.00  &    0.86  &    0.92    &     7\\
     linke    &   1.00   &   1.00     & 1.00    &     7\\
       spd   &    0.80   &   0.50    &  0.62     &    8\\
    \hline
    avg / total  &     0.88   &   0.86   &   0.86  &      36\\
\end{tabular}
\end{center}

@@

\subsection{Validation}
\paragraph{Differentiability of policy issues}
One reason why some statements are difficult to classify could be that some issues are more difficult to differentiate than others. To test this a separate suite of experiments was run to train and test the prediction performance of the text classifiers models described in \autoref{sec:model}. As there was no out-of-domain evaluation set available in this setting only the evaluation error on in-domain data is reported. Note however that in this experiment too the evaluation data was never seen by any model during training time.
In \autoref{tab:results_avg_political_view} results for the best and worst classes, in terms of predictability, are listed along with the average performance metrics on all classes.
Precision and recall values of close to 0.5 on average can be considered rather high considering the large number of labels. Results are not surprising, but are equal to those by human coders. \\

\begin{table}
\caption{
\label{tab:results_avg_political_view}
Classification of 56 political views, see \autoref{sec:data}.
}
\begin{center}
\footnotesize
\begin{tabular}{lcrrrr}
code & meaning  &      prec    &recall & N\\
\hline\hline
%
       501  & environmentalism + &      0.62   &   0.61 &      165\\
        202 &   democracy + &   0.58  &    0.55   &      122\\
        701    & labour +&  0.57  &    0.54   &   129\\
                201    &freedom/human rights +  & 0.58   &   0.54   &      159\\
         106   & peace + & 0.52&      0.57   &       21\\
\dots\\
        302     &centralism + & 0.25     & 0.20  &          10\\
        401    &  free enterprise + &0.20    &  0.19    &       52\\
        505    &welfare - &  0.13   &   0.14   &      14\\
        409    & keynesian demand +&  0.14  &      0.13   &      8\\
            0    & undefined &  0.09  &    0.12   &    17\\
            \hline
avg / total  &  &  0.47    &  0.46 &    2946\\
%
\end{tabular}
\end{center}
\end{table}

\paragraph{Policy change}
Another explanation for missclassification could lie in the fact that parties change their policy positions. In order to investigate this confusion matrices were extracted for the predictions on the out-of-domain evaluation data for sentence level predictions (see \autoref{tab:confusion}) as well as topic level predictions (see \autoref{tab:confusion_topic}). On the topic level results for the 17th legislative period are pretty good except for the SPD. In the 18th legislative period predictions are very bad for the Green party. One explanation for this missclassification could lie in policy changes. For example, the Green party has devoted large parts of its manifesto to renewable and against nuclear energy. After Fukushima the other parties adopted these positions. \\

% Die könnten wir vielleicht rausnehmen.
% Confusion matrix sentence level
\begin{table}[t]\label{tab:conf_mat_four_class}
\caption{\label{tab:confusion} {\bf Confusion matrices (sentence level)} for predictions on evaluation data (party manifestos); classifiers were trained on parliament speeches for the 17th legislative period; the most prominent effect is the high likelihood for a party to be taken as the strongest, governing party, cdu/csu. This can be interpreted as a change in policies of the conservative party cdu/csu towards the policies of the green party.}
\begin{tabular}{lccccccc}
 \multicolumn{8}{c}{\bf Party Manifestos}\\
 \multicolumn{8}{c}{\bf 17th Bundestag}\\
 \\
&&& \multicolumn{5}{c}{Predicted}\\
&&& cducsu & fdp& gruene& linke& spd\\
\hline
\multirow{5}{*}{\rotatebox{90}{\pbox{3cm}{\centering True}}}& &cducsu &1186 &289& 178& 198& 179\\
&&fdp &882& 658& 236& 329& 214\\
&&gruene &1174& 404& 764& 941& 464\\
&&linke &388& 92& 214& 806& 201\\
&&spd &999& 268& 240& 398& 373\\
\end{tabular}
\end{table}

% Confusion matrix topic level
\begin{table}[t]\label{tab:conf_mat_four_class}
\caption{\label{tab:confusion_topic} {\bf Confusion matrices (topic level)} for predictions on evaluation data (party manifestos) for classifiers trained on parliament speeches for the 17th legislative period (left) and 18th legislative period (right).}
\begin{tabular}{lccccccc}
 \multicolumn{8}{c}{\bf Party Manifestos}\\
 \multicolumn{8}{c}{\bf 17th Bundestag}\\
 \\
&&& \multicolumn{5}{c}{Predicted}\\
&&& cducsu & fdp& gruene& linke& spd\\
\hline
\multirow{5}{*}{\rotatebox{90}{\pbox{3cm}{\centering True}}} &&cducsu &7& 0& 0& 0& 0\\
&&fdp&0& 7& 0& 0& 0\\
&&gruene&0& 0& 6& 0& 1\\
&&linke&0& 0& 0& 7& 0\\
&&spd&4& 0& 0& 0& 4\\
\end{tabular}
\end{table}
